Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2019-07-02 write-up/fda-recurrent.tex
Fixing up paper: working on proofs of technical results

\newcommand{\sam}[1]{{\color{blue}{#1}}}
\newcommand{\walt}[1]{\textcolor{red}{[WD:\ #1]}}

@@

\usepackage{soul,color, mathtools}

\def\pr{\mathop{\text{pr}}\nolimits}

@@

The goal of this paper is two-fold: (1) to discuss the appropriate
choice of statistical model for joint high-frequency sensor data and
the recurrent event data, and (2) to construct a robust,
easy-to-implement method for parameter estimation and inference. 
For (1), we discuss an important issue regarding measurement-error
models when paired with recurrent event outcomes.
For (2), we introduce a random subsampling procedure that has
several benefits.  First, the resulting inference is unbiased;
however, there is a computation-efficiency trade-off. In particular, a
higher sampling rate can decrease variance at the cost of increased
computation.  We show via simulations that the benefits of incredibly
high sampling rates is often negligible, as the contribution to the
variation is small in relation to to variation in the underlying
stochastic processes. Second, optimal estimating equations are derived
that depend on the instantaneous rate of the recurrent event process,
which increases efficiency of the subsampling procedure. 
Finally, implementation can leverage existing, standard software for
functional data analysis and logistic regression, leading to fast
adoption by domain scientists.

@@

Suppose $n$ subjects are independently sampled with observed event
times~$\bfT_{i} = \{ T_{i,1}, \ldots, T_{i,k_i}\}$ over some observation
window $[0, \tau_i]$ for each subject $i = 1,\ldots, n$.  Assume the
event times are ordered, i.e., $T_{i,j} < T_{i,j^\prime}$ for $j <
j^\prime$. The window length, $\tau_i$, is the censoring time and is
assumed independent of the event process. Let~$N_{i} (t)$ denote the
associated counting process of $\bfT_{i}$; that is, $N_i (t) =
\sum_{j=1}^{k_i} 1 [ T_{i,j} < t ]$.  
In this section, we assume a single-dimensional health process~$\bfx_i
= \{ x_i (s) \}_{0 < s < \tau_i}$ for each participant is measured at
a dense grid of time points.  Accelerometer, for example, is measured
at a rate of 32Hz (i.e., $32$ times per second). Electrodermal
activity (EDA), on the other hand, is measured at a rate of 4Hz (i.e.,
4 times per second).  Given the high frequency nature of the sensor 
measurements, this paper assumes the process is measured continuously.

@@

An alternative to~\eqref{eq:hazardlinear} would be to construct
features from the sensor data history~$f_t ( H_{i,t}^{X}) \in
\mathbb{R}^q$ and incorporated these features in the place of the
final term. Our current approach builds linear features of~$H_{i,t}^X$
directly from the integrated history, avoiding the feature
construction problem -- a highly nontrivial issue for high frequency
time-series data.  The main caveat is the additional
parameter~$\Delta$; however, we show that so long as the
estimated~$\hat \Delta$ exceeds~$\Delta$, then resulting estimation is
unbiased albeit at a loss of efficiency.  Moreover, sensitivity
analysis can be peformed to determine how choice of $\hat \Delta$
affects inference.  One limitation of the approach presented here is
that only fully parametric hazard models may be fit to the data.
However, a spline model for the log baseline hazard affords sufficient
model flexibility.

\subsection{Measurement-error models with event processes}

One potential criticism for~\eqref{eq:hazard} is that the health
process may be measured with error.  
A common mathematical strategy for joint models is to consider an
unobservable, latent process~$\eta_i$ such that~$\bfT_i \indep x_i
\given \eta_i$. Take~$\eta_i$ to be a zero-mean Gaussian process with 
\begin{equation}\label{eq:jm}
x_i(t) = \eta_i (t) + \epsilon_i (t),\quad \text{ and } \log h_i (t
\given \eta ) = \log h_0 (t) + g_t \left( H_{i,t}^N \right)^{\prime}
\alpha + \int_{t-\Delta}^t \eta_i (s) \beta (s) ds
\end{equation}
where~$\epsilon (t)$ is a white-noise measurement error. Thus,
$\eta_i (t)$ is the ``true and unobserved value of the longitudinal
outcome'' \citep[Sec. 2.1, pp.3]{Rizopoulos2010}. The conditional intensity
function given~$\bfx_i$ is given by
\[
\pr \left ( T_{i,j} = t \Given \H_{i,\tau_i}^X, \{ T_{i,j^\prime} < t \}_{j^\prime <
  j}, T_{i,j} \geq t \right) = E \left( h_i (t \given \eta) \exp
\left( \int_{T_{i,j-1}}^t h_i (\eta (s)) ds \right) \given \bfx_i
\right).
\]
Therefore, it appears that the conditional hazard function depends 
not only on past $x$-values, but also on future $x$-values.
Therefore, the processes~$\bfx$ and~$\bfT$ do not satisfy
\emph{independent evolution}~\citep{DempseyPMCC2}.
This is quite unnatural, as~\eqref{eq:jm} suggests current risk
depends on future values of the sensor process.
To alleviate this issue, we employ the strategy of taking~$\epsilon
(t) \equiv 0$ and treating~$\bfx_i$ as a mean-zero Gaussian process.
The proposed inferential procedure only relies on the Gaussian
assumption when~$\bfx$ is observed with missingness (see
Section~\ref{section:missingdata}).

@@

To solve the computational challenge we employ a point-process
subsampling design to obtain unbiased estimates of the derivative of
the cumulative hazards for each subject. The subsampling procedure
treats the collected sensor data as a set of \emph{potential
  observations}. Suppose covariate information is sampled at times
drawn from an independent inhomogeneous Poisson point process with
known intensity~$\pi_i (t)$. At a subsampled time~$t$, the
\emph{windowed covariate history} $\{ x_i (t-s)\}_{0 \leq s \leq
  \Delta}$ and counting process features~$g_t (H_{i,t}^N)$ are
observed. Optimal choice of~$\pi_i (t)$ is beyond the scope of this
paper; however, the literature suggests setting proportional to the
hazard function~$h_i (t; \theta)$.

An estimator is design-unbiased if its expectation is equal to that
parameter under the probability distribution induced by the sampling
design~\citep{Cassel1977}. Let~$D_i \subset [0,t_i]$ denote the set of
subsampled points.  Note, by design, this set is distinct from the set
of event times, i.e.,~$\bfT_i \cap D_i = \emptyset$.  Under
subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson
estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in D_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (s) }
\]
An alternative design-unbiased estimator of the derivative of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup D_i)}
\frac{ h^{(1)}_i ( t; \theta ) }{ \pi_i (s) + h_i ( t ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested
by~\cite{Waagepetersen2008}.  This estimator depends on the
superposition of the event and subsampling processes.
Proposition~\ref{prop:optimal} shows the estimator for~$\theta$
associated with using~\eqref{eq:WPest} is the most efficient within a
suitable class of estimators for the derivative of the cumulative
hazard function (including the Horvitz-Thompson estimator).
Therefore, we restrict our attention to~\eqref{eq:WPest} for the
remainder of this paper. Letting
\[
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\]
the resulting estimating equations can be written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( t; \theta)}  - \sum_{u \in D_i} w_i(u;
  \theta) \frac{h_i^{(1)} (t; \theta)}{ \pi_i (t) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built
via plug-in of the design-unbiased estimator of the derivative of the
cumulative hazard given in~\eqref{eq:WPest}.

@@

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0
\leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup D_i$.
We use the sandiwch smoother~\citep{Xiao2013} to estimate the
mean~$\mu(s,t) = \E [ X(t,s)]$.
Alternative bivariate smoothers exist, such as the kernel-based local
linear smoother~\citep{Hastie2009} bivariate tensor product
splines~\cite{Wood2006} and the bivar
The sandwich smoother was chosen for its computational efficiency and
estimation accuracy.
We then define~$\tilde X(s,t) = X(s,t) - \hat \mu(s,t)$ to be the
mean-zero process at each time~$t \in $.

Next,

Use of the marginal covariance~\cite{Park2018} reduces allows us to avoid
computationally prohibitive computations.

\begin{itemize}
\item {\bf Step 1}: Estimate~$\hat \mu(s,t)$ via sandwich smoother for
  all~$s \in [0,\Delta]$ and~$t \in \cup_{i=1}^n \left( D_i \cup N_i
  \right)$. Define~$\tilde X (s,t) = X(s,t) - \hat \mu(s,t)$ to be a
  mean-zero process for each~$t$.
\item {\bf Step 2}: Compute the pooled sample covariance
\[
\hat \Xi (s , s^\prime) = 
\left( \sum_{i=1}^n \left | N_i \cup D_i \right |\right)^{-1}
\left( \sum_{i=1}^n \sum_{t \in N_i \cup D_i} \tilde X (s,t) \tilde X
  (s^\prime, t) \right).
\]
Compute at a fine set of grid points. This is an estimate of 
\[
\Xi (s, s^\prime) = \int_{\mathcal{T}} c( (s,T), (s^\prime, T) ) g(T) dT
\]
where~$g(T)$ is a superposition of the event process and the
subsampling process.
\item {\bf Step 3}: Take spectral decomposition~$\{ \hat \phi_k (s),
  \hat \lambda_k \}_{k \geq 1}$. Choose~$K_x \geq K_b$ large enough.
\item {\bf Step 4}: Compute
\[
\hat c_{i,k} (t) = \int_{t-\Delta}^t \tilde X_i (s,t) \hat \phi_k (s) ds
\]
Then
\[
X(s,t) \approx \hat \mu(s,t) + \sum_{k=1}^{K_x} \hat c_{i,k} (t)
\phi_k (s).
\]
\item {\bf Step 5}: Plugging both the spline and KL decomposition we
  have
\begin{align*}
\int_{t-\Delta}^t X(s,t) \beta(s) ds 
  &= \int_{t-\Delta}^t \left[ \hat \mu(s,t) + \sum_{k=1}^{K_x}
    c_{i,t,k} \phi_k (s) \right] \times \left[ \sum_{k^\prime=1}^{K_b}
    \beta_k \psi_k (s) \right] \\
  &= [ M_{i,t}^\prime + C_{i,t}^\prime J_{\phi, \psi} ] \beta
\end{align*}
where~$M_{i,t}$ is a vector with~$i$th entry equal
to~$\int_{t-\Delta}^t \hat \mu(s,t) \psi_k (s) ds$, $C_{i,t} =
(c_{i,t,1}, \ldots, c_{i,t,K_x}) \in \mathbb{R}^{K_x \times 1}$,
$\beta = (\beta_1, \ldots, \beta_{K_b}) \in \mathbb{R}^{K_b \times 1}$
and~$J_{\phi, \psi} \in \mathbb{R}^{K_x \times K_b}$ where the~$(i,j)$
entry is given by
\[
\int_{t-\Delta}^t \phi_{i} (s) \psi_{j} (s) ds.
\]
\item {\bf Step 6}: The approximate score equations
\begin{equation}
\label{eq:weighted_esteq}
\hat U_n (\theta) = \sum_{i=1}^n \left \{ \sum_{t \in N_i} w_{i} (t; \theta) W_{i,t} -
    \sum_{u \in D_i} (1 - w_i (u; \theta)) W_{i,u} \right \} 
\end{equation}
where
\begin{equation}
\label{eq:waage_weight}
w_i (t; \theta) = \frac{h_i(t;\theta)}{\pi_i (t) + h_i (t;\theta)}.
\end{equation}
Under weights~\eqref{eq:waage_weight} and log-linear intensity
function,~\eqref{eq:weighted_esteq} is the logistic score function for
the binary variable~$Y_i(u)$ for $u \in N_i \cup D_i$ and~$i \in [n]$
where~$Y_i(u) = 1[u \in N_i]$. To see this, we rewrite $\hat U_n
(\theta)$ as 
\[
  - \sum_{i=1}^n \sum_{t \in N_i \cup D_i}
    \left[ {\bf 1} [t \in D_i]  - w_{i} (t; \theta) \right] W_{i,t} 
= \sum_{i=1}^n \sum_{t \in N_i \cup D_i}
    \left[ {\bf 1} [t \in D_i]  - \frac{1}{1 + \exp\left( - (
          \tilde W_{i,t}^\top \theta + \log (\pi_i (t) ) ) \right)}
    \right] \tilde W_{i,t}.
\]
where~$\tilde W_{i,t} = - W_{i,t}$.
This is equivalent to
\[
\sum_{i=1}^n y_i \log p (x_i)
\]
\end{itemize}

\begin{thm}[Asymptotic theory]
For large~$n$, the estimate~$\hat \theta_n$ is approximately normal
\[
\sqrt(n) (\hat \theta_n - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is a random 
\end{thm}
An estimator for~$\Xi(\theta)$ is
\[
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{u \in N_i \cup D_i} 
  w_i(u;\theta) (1-w_i(u;\theta)) \left[ \frac{h_i^{(1)}(u;
      \theta)}{h_i (u; \theta)} \right] \times  \left [ \frac{h_i^{(1)} (u;\theta)}{h_i(u; \theta)} \right]^\top
\]
For the Cox model, the sampling-unbiased estimator for~$\hat
\Xi(\theta)$ is equivalent to the Fisher information for the
previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process,
standard logistic regression software can be used to fit the recurrent
event model by specifying an offset equal to~$\log (\pi_i (s))$.

\begin{prop}[Optimality of proposed approach]
\label{prop:optimal}
If the event process is an inhomogeneous Poisson point process with
intensity~$h(t; \theta)$ and subsampling occurs via an independent,
inhomogeneous Poisson point process with intensity~$\pi (t)$,
then~$\hat U_n (\theta)$ are optimal estimating functions in a class
of weighted estimating functions given by~\eqref{eq:weighted_esteq}
with replacing~\eqref{} by any weight function~$w_i (t; \theta)$. This
class includes the Horvitz-Thompson estimator under~$w(s; \theta)
= 1$.
\end{prop}

Proposition~\ref{} follows immediately from~\cite{Rathbun2013}.
This ensures the only loss of statistical efficiency is related to the 
subsampling procedure and not the associated methodology.

@@

\begin{assumption}[Event process assumptions]

\end{assumption}

\begin{assumption}[Covariance assumptions]

\end{assumption}

\begin{assumption}[Functional data assumptions]

\end{assumption}

\begin{thm}[No reconstruction error]

@@

\begin{thm}[Reconstruction error but~$\hat \Delta > \Delta$]
\label{eq:biaseddown}
\end{thm}

Theorem~\ref{eq:}

\begin{thm}[When~$\hat \Delta < \Delta$]

@@

>>>>

%\newcommand{\sam}[1]{{\color{blue}{#1}}}
\newcommand{\walt}[1]{\textcolor{blue}{[WD:\ #1]}}

@@

\usepackage{soul,color, mathtools, fixmath}
\usepackage{multirow}
\def\pr{\mathop{\text{pr}}\nolimits}

@@

The goal of this paper is two-fold: (1) to discuss the appropriate
choice of statistical model for joint high-frequency sensor data and
the recurrent event data, and (2) to construct a robust,
easy-to-implement method for parameter estimation and inference. 
For (1), we discuss an important issue regarding measurement-error
models when paired with recurrent event outcomes.
For (2), we introduce a random subsampling procedure that has
several benefits.  First, the resulting inference is unbiased;
however, there is a computation-efficiency trade-off. In particular, a
higher sampling rate can decrease estimator variance at the cost of
increased computation.  We show via simulations that the benefits of
incredibly high sampling rates is often negligible, as the
contribution to the variation is small in relation to to variation in
the underlying stochastic processes. Second, derived estimating
equations are optimal, implying loss of statistical efficiency is only
due to subsampling procedure and not derived methodology.  Finally,
implementation can leverage existing, standard software for functional
data analysis and logistic regression, leading to fast adoption by
domain scientists.

@@

Suppose $n$ subjects are independently sampled with observed event
times~$\bfT_{i} = \{ T_{i,1}, \ldots, T_{i,k_i}\}$ over some observation
window $[0, \tau_i]$ for each subject $i = 1,\ldots, n$.  Assume the
event times are ordered, i.e., $T_{i,j} < T_{i,j^\prime}$ for $j <
j^\prime$. The observation window length, $\tau_i$, is the censoring
time and is assumed independent of the event process. Let~$N_{i} (t)$
denote the associated counting process of $\bfT_{i}$; that is, $N_i
(t) = \sum_{j=1}^{k_i} 1 [ T_{i,j} < t ]$.  
In this section, we assume a single-dimensional health process~$\bfx_i
= \{ x_i (s) \}_{0 < s < \tau_i}$ for each participant is measured at
a dense grid of time points.  Accelerometer, for example, is measured
at a rate of 32Hz (i.e., $32$ times per second). Electrodermal
activity (EDA), on the other hand, is measured at a rate of 4Hz (i.e.,
4 times per second).  Given the high frequency nature of sensor data,
this paper assumes the process is measured continuously.

@@

An alternative to~\eqref{eq:hazardlinear} would be to construct
features from the sensor data history~$f_t ( H_{i,t}^{X}) \in
\mathbb{R}^q$ and incorporated these features in the place of the
final term. Our current approach builds linear features of~$H_{i,t}^X$
directly from the integrated history, avoiding the feature
construction problem -- a highly nontrivial issue for high frequency
time-series data.  The main caveat is the additional
parameter~$\Delta$; however, as long as the estimated~$\hat \Delta$
exceeds~$\Delta$, then resulting estimation is unbiased albeit at a
loss of efficiency.  Moreover, sensitivity analysis can be peformed to
determine how choice of $\hat \Delta$ affects inference.  One
limitation of the approach presented here is that only fully
parametric hazard models may be fit to the data. However, a spline
model for the log baseline hazard affords sufficient model
flexibility.

\subsection{Measurement-error models with event processes}
\label{section:memproblems}

One potential criticism for~\eqref{eq:hazard} is that the health
process may be measured with error.  
A common mathematical strategy for joint models is to consider an
unobservable, latent process~$\eta_i$ such that~$\bfT_i \indep x_i
\given \eta_i$. Take~$\eta_i$ to be a zero-mean Gaussian process with 
\begin{equation}\label{eq:jm}
x_i(t) = \eta_i (t) + \epsilon_i (t),\quad \text{ and } \log h_i (t
\given \eta ) = \log h_0 (t) + g_t \left( H_{i,t}^N \right)^{\prime}
\alpha + \int_{t-\Delta}^t \eta_i (s) \beta (s) ds
\end{equation}
where~$\epsilon (t)$ is a white-noise measurement error. Thus,
$\eta_i (t)$ is the ``true and unobserved value of the longitudinal
outcome'' \citep[Sec. 2.1, pp.3]{Rizopoulos2010}. The conditional
density function of the event process~$\bfT_i$ given~$\bfx_i$ is 
\[
\pr \left ( \bfT_i = \bft \Given \H_{i,\tau_i}^X \right) = E \left(
  \prod_{j=1}^{k_i} h_i (t_j \given \eta) \exp \left(
    \int_{0}^{\tau_i} h_i (s \given \eta ) ds \right) \given \bfx_i \right).
\]
Therefore, the conditional density function depends not only on past
$x$-values, but also on future $x$-values.
That is, the processes~$\bfx$ and~$\bfT$ do not satisfy
\emph{independent evolution}~\citep{DempseyPMCC2}. 
This is quite unnatural, as~\eqref{eq:jm} suggests the instantaneous
risk of an event at time~$t$ depends on future values of the sensor
process. To alleviate this issue, we employ the strategy of
taking~$\epsilon (t) \equiv 0$ and treating~$\bfx_i$ as a mean-zero
Gaussian process. The proposed inferential procedure only relies on
the Gaussian assumption when~$\bfx$ is observed with missingness (see 
Section~\ref{section:missingdata}).

@@

To solve the computational challenge we employ a point-process
subsampling design to obtain unbiased estimates of the derivative of
the cumulative hazards for each subject. The subsampling procedure
treats the collected sensor data as a set of \emph{potential
  observations}. Suppose covariate information is sampled at times
drawn from an independent inhomogeneous Poisson point process with
known intensity~$\pi_i (t)$. At a subsampled time~$t$, the
\emph{windowed covariate history} $\{ x_i (t-s)\}_{0 \leq s \leq
  \Delta}$ and counting process features~$g_t (H_{i,t}^N)$ are
observed. Optimal choice of~$\pi_i (t)$ is beyond the scope of this
paper; however, simulation studies have suggested setting proportional
to the hazard function~$h_i (t; \theta)$.

An estimator is design-unbiased if its expectation is equal to that
parameter under the probability distribution induced by the sampling
design~\citep{Cassel1977}. Let~$D_i \subset [0,t_i]$ denote the set of
subsampled points.  Note, by design, this set is distinct from the set
of event times, i.e.,~$\bfT_i \cap D_i = \emptyset$.  Under
subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson
estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in D_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (s) }
\]
An alternative design-unbiased estimator of the derivative of the
cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup D_i)}
\frac{ h^{(1)}_i ( t; \theta ) }{ \pi_i (s) + h_i ( t ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested
by~\cite{Waagepetersen2008}.  This estimator depends on the
superposition of the event and subsampling processes.
Proposition~\ref{prop:optimal} shows the estimator for~$\theta$
associated with using~\eqref{eq:WPest} is the most efficient within a
suitable class of estimators for the derivative of the cumulative
hazard function (including the Horvitz-Thompson estimator).
Therefore, we restrict our attention to~\eqref{eq:WPest} for the
remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting estimating equations can be written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( t; \theta)}  - \sum_{u \in D_i} w_i(u;
  \theta) \frac{h_i^{(1)} (t; \theta)}{ \pi_i (t) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built
via plug-in of the design-unbiased estimator of the derivative of the
cumulative hazard given in~\eqref{eq:WPest}.

@@

\subsection{Estimation of the windowed covariate history}

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0
\leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup D_i$.
We use the sandwich smoother~\citep{Xiao2013} to estimate the
mean~$\mu(s,t) = \E [ X(t,s)]$.
Alternative bivariate smoothers exist, such as the kernel-based local
linear smoother~\citep{Hastie2009} bivariate tensor product
splines~\citep{Wood2006} and the bivariate penalized spline
smoother~\citep{MarxEilers2005}.
The sandwich smoother was chosen for its computational efficiency and
estimation accuracy.
We then define~$\tilde X(s,t) = X(s,t) - \hat \mu(s,t)$ to be the
mean-zero process at each time~$t \in \bfT_i \cup D_i$.

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma (s, s^\prime) = \int_{0}^\tau c( (s,T), (s^\prime, T) ) g(T) dt.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c((s,t), (s^\prime,t))$ is
the covariance function of the windowed covariate history~$X(t,\cdot)$
and $g(T)$ is the joint sampling density of event and subsampled
timepoints. 
Estimation of~$\Sigma$ occurs in two steps. 
First, the pooled sample covariance is calculated at a set of grid
points:
\[
\tilde \Sigma (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | N_i 
    \cup D_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in N_i
    \cup D_i} \tilde X (s_r,t) \tilde X (s_{r^\prime}, t) \right).
\]
Due to our concern over independent evolution as discussed in
section~\ref{section:memproblems}, we do not assume that each
observation is observed with white noise and therefore the diagonal
elements of~$\hat \Sigma$ are not inflated. Second, the
estimator~$\hat \Sigma$ is further smoothed again using the sandwich 
smoother~\citep{Xiao2013}. 
Note~\cite{Park2018} smooth the off-diagonal elements, while here we
smooth the entire pooled sample covariance matrix. All negative
eigenvalues are set to zero to ensure positive semi-definiteness.
The result is used as an estimator~$\hat \Sigma$ for the pooled
covariance $\Sigma$.

Next, we take the spectral decomposition of the estimated covariance
function; let~$\{ \hat \psi_k (s),\hat \lambda_k \}_{k \geq 1}$ be the
resulting sequence of eigenfunctions and eigenvalues. 
The key benefit of the marginal covariance approach is that it allows
us to compute a single, time-invariant basis expansion; this reduces
the computational burden by avoiding the three dimensional covariance
function (i.e., covariance depends on~$t$) and associated spectral
decomposition in methods considered by \cite{ChenMuller2012}.
Using the Karhunen-Lo{\`e}ve decomposition, we can represent~$X(t,s)$
by
\[
X(s,t) = \hat \mu(s,t) + \sum_{k=1}^{\infty} \hat c_{i,k} (t) \hat
\psi_k (s) \approx \hat \mu(s,t) + {\bf c}_{i} (t)^\top \mathbold{\hat
  \psi} (s)
\]
where $\hat c_{i,k} (t) = \int_{t-\Delta}^t \tilde X_i (s,t) \hat
\psi_k (s) ds$, ${\bf c}_i (t) = (c_{i,1} (t), \ldots, c_{i,K_x}
(t))^\top$, $\mathbold{\hat \psi} (s) = (\hat \psi_1 (s), \ldots, \hat
\psi_{K_x} (s))^\top$, and~$K_x < \infty$ is the truncation level of
the infinite expansion. Following~\cite{Goldsmith2011}, we set~$K_x$
to satisfy identifiability constraints (see Section~\ref{section:beta}
for details).

\subsection{Estimation of~$\beta(s)$}
\label{section:beta}

The next step of our method is modeling~$\beta(s)$. Here, we leverage
ideas from the penalized spline literature~\citep{Ruppert2003,
  Wood2006book}. Let~$\mathbold{\phi} (s) = \{ \phi_1 (s), \ldots,
\phi_{K_b} (x) \}$ be a spline basis such that~$\beta(s) =
\sum_{j=1}^{K_b} b_j \phi_{j} (s) = \mathbold{\phi} (t) {\bf b}$
where~${\bf b} = [b_1, \ldots, b_{K_b}]^{\top}$. Thus, the integral
in~\eqref{eq:hazardlinear} can be restated as 
\begin{align*}
\int_{t-\Delta}^t X(s,t) \beta(s) ds 
  &= \int_{t-\Delta}^t \left[ \hat \mu(s,t) + {\bf c} (t)^\top
    \mathbold{\hat \psi} (s) \right] \times \left[
    \mathbold{\phi} (s) {\bf b} \right] \\
  &= [ M_{i,t}^\prime + C_{i,t}^\prime J_{\psi, \phi} ] {\bf b}
\end{align*}
where~$M_{i} (t) = (M_{i,1} (t), \ldots, M_{i,K_b} (t))$, $M_{i,j} (t)
= \int_{t-\Delta}^t \hat \mu (s,t) \phi_j (s)$, and~$J_{\psi, \phi}$ is a
$K_x \times K_b$ dimensional matrix with the~$(k,l)$th entry is equal
to~$\int_{0}^\Delta \psi_k (s) \phi_l (s)
ds$~\citep{RamsaySilverman2005}.

Given the basis for~$\beta(t)$, the model depends on choice of
both~$K_b$ and~$K_x$.  We follow~\cite{Ruppert2002} by choosing $K_b$
large enough to prevent undersmoothing and~$K_x \geq K_b$ to satisfy
identifability constraints.
In this paper, we follow the simple rule of thumb and set~$K_b = K_x =
35$.
As long as the choices of~$K_x$ and $K_b$ are large enough, their
impact on estimation is typically negligible. 
We formalize this statement in Theorem~\ref{thm:truncation}.
Note the choice of~$K_b$ large may lead to overfitting the data.
We therefore follow~\cite{GoldSmith2015} and induce smoothing by
assuming that~$\{ b_k \}_{k=3}^{K_b} \sim N(0, \sigma_b^2 I)$.
Below, we will exploit a connection between~\eqref{eq:approxscore}
and score equations for a logistic regression model.  Before moving
on, we introduce notation:
\begin{equation*}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) \approx
\exp \left( Z_{i,t}^\top \gamma + g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + M_{i,t}^\top {\bf b} + C_{i,t}^\top J_{\phi, \psi} {\bf b} \right)
= \exp \left( W_{i,t}^\top \theta \right),
\end{equation*}
where~$\theta = (\gamma, \alpha, {\bf b})$ and~$\exp ( Z_{i,t}^\top
\gamma) =: h_0 (t)$ is the parametrized baseline intensity function.

\subsection{Connection to logistic score functions}

We next establish a connection between the above approximate score 
equations and the score equations for a logistic regression model.
All proofs are provided in the supplementary materials.
We can then exploit this connection to allow the model to be fit
robustly using standard mixed effects software~\citep{Ruppert2002,
  McCulloch2001}.

\begin{lemma} \normalfont
\label{lemma:logistic}
Under weights~\eqref{eq:waage_weights} and the log-linear intensity
function~\eqref{eq:hazardlinear}, the approximate score
function~\eqref{eq:approxscore} is equivalent to 
\[
\sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i } \left[ {\bf 1}[t \in D_i]
  - \frac{1}{1 + \exp \left[- \left( \tilde{W}_{i,t}^\top \theta +
        \log \pi_i (t) \right) \right]} \right] \tilde{W}_{i,t} 
\]
where~$\tilde W_{i,t} = -W_{i,t}$. This is the score function for
logistic regression with binary response~$Y_i(t)$ for $t \in N_i \cup
D_i$ and~$i \in [n]$ where~$Y_i(t) = 1[t \in N_i]$, offset~$\log \pi_i
(t)$, and covariates~$\tilde W_{i,t}$. 
\end{lemma}

In this section, we make the following strong assumption.

\begin{assumption} \normalfont
\label{assumption:truncation}
Both the functional covariate~$\{ X(t,s) \}_{0\leq s \leq \Delta}$ and
the functional coefficient~$\beta(t)$ sit in the span of their
respective truncated basis expansions. That is, there
exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$
and~$X(s,t) = \hat \mu (s,t) + {\bf c}_i (t)^\top \mathbold{\hat \psi}
(s)$.
\end{assumption}

Assumption~\ref{assumption:truncation} allows us to ignore the impact
of truncation and present Lemma~\ref{lemma:simpleasym} which gives a
simple asymptotic theory in this particular case.

\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:truncation} and~$\Delta$ known, for
large~$n$ the estimate~$\hat \theta_n$ is approximately normal
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
An estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i} 
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator
for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the
previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process,
standard logistic regression software can be used to fit the recurrent
event model by specifying an offset equal to~$\log \pi_i (t)$.
Not only this, Lemma~\ref{prop:optimal} shows
weights~\eqref{eq:waage_weights} are optimal within a particular class
of weighted estimating equations.

\begin{prop} \normalfont
\label{prop:optimal}
If the event process is an inhomogeneous Poisson point process with
intensity~$h(t; \theta)$ and subsampling occurs via an independent,
inhomogeneous Poisson point process with intensity~$\pi (t)$,
then~$\hat U_n (\theta)$ are optimal estimating functions (i.e., most
efficient) in the class of weighted estimating functions given
by~\eqref{eq:approxscore} replacing~\eqref{eq:waage_weights} by any
weight function~$w_i (t; \theta)$. This class includes the
Horvitz-Thompson estimator under~$w(s; \theta) = 1$.
\end{prop}

\noindent Proposition~\ref{prop:optimal} ensures the only loss of
statistical efficiency is due to subsampling and not using a
suboptimal estimation procedure given subsampling.

@@

\begin{assumption}[Event process assumptions]
\begin{itemize}
\item The subsampling rate~$\pi_i (t) \in [L_\pi, U_\pi]$ 
\item There exists a nonnegative matrix
\item There exists 
\end{itemize}
\end{assumption}

\begin{assumption}[Covariance assumptions]
\begin{itemize}
\item 
\end{itemize}
\end{assumption}

\begin{assumption}[Functional data assumptions]
M1 -- M4 in 
\end{assumption}

\begin{thm}[No reconstruction error]
\label{thm:truncation}
Under regularity conditions A-E in Andersen et al. 1993, 
\[
\frac{n (\hat \theta - \theta) \Xi (\theta) (\hat \theta - \theta) -
  K_n+1}{\sqrt{2 (K_n + 1) }} \to N(0,1).
\]

@@

\begin{thm}[Reconstruction error OR $\hat \Delta > \Delta$]
\label{thm:biaseddown}

@@

\subsection{Re-using estimates under additional sampling}
